Verification of OpenSSL's SHA3 Keccak function using the Software Analysis Workbench

### 2021-09-10 paper/amended.tex
Update amended.tex

\section{Amended Cryptol}\label{sec:amended}
TODO:

Imperative spec, functional language, Cryptol's looping syntax can be cryptic.

Rational and descriptions of utility functions, examples of utility functions.

for-each == javascript map, while loop == javascript reduce.

Initially it appeared that visual inspection with Cryptol’s standard syntax was subpar.
Cryptol’s functional folds and list comprehensions do not align easily with the imperative 
and array-based NIST standard. BY leveraging Cryptol’s versatility, however, we produced a couple 
utility functions to visually mimic the sequential nature of the specification.

The phrase “for all” is quite common in the NIST document, and it implies a looping structure. 
The following is the NIST definition of the pi function:
\begin{verbatim}
  1. For all triples (x,y,z) such that 
    0 <= x < 5, 0 <= y < 5, and 0 <= z < w,
    let A'[x, y, z] = A[(x + 3y) mod 5, x, z].
  2. Return A'.
\end{verbatim}

Using Cryptol's list comprehensions, we find the following syntax for the pi function:
\begin{verbatim}
  type COL = 5 
  type W = 64
  LIST4 = [0..COL-1]
  LIST63 = [0..W-1]
  type STATE_ARR = [COL][COL][W]
  pi : STATE_ARR -> STATE_ARR
  pi a = [[[a @x @((x + 3y) % COL) @z 
             | z <- LIST63] 
           | y <- LIST4] 
         | x <- LIST4]
\end{verbatim}

@@

\begin{verbatim}
  type COL = 5 
  type W = 64
  type STATE_ARR = [COL][COL][W]
  LIST4 = [0..COL-1]
  LIST63 = [0..W-1]
  pi : STATE_ARR -> STATE_ARR
  pi a = for LIST4 (\y ->
           for LIST4 (\x -> 
             for LIST4 (\z -> 
               a @x @((x + 3*y) % `COL) @z)))
\end{verbatim}

@@

\begin{verbatim}
  for : {n, a, b} [n]a -> (a -> b) -> [n]b
  for vals loop = [loop index | index <- vals]
\end{verbatim}

@@

\begin{verbatim}
  1. If t mod 255 = 0, return 1.
  2. Let R = 10000000.
  3. For i from 1 to t mod 255, let:
    a. R = 0 || R;
    b. R[0] = R[0] ^ R[8];
    c. R[4] = R[4] ^ R[8];
    d. R[5] = R[5] ^ R[8];
    e. R[6] = R[6] ^ R[8];
    f. R =Trunc8[R].
  4. Return R[0].
\end{verbatim}

@@

\begin{verbatim}
  rc : [64] -> Bit
  rc t = if (t % 255) == 0 
    then True 
      else (rList !0).R @0 where
        rList = [{i = 1, R = 128:[8]}] # [{
          i = r @i + 1, 
          R = if i <= t % 255
            then (take`{8} ([
              (r@0)^(r@8),
              r@1,
              r@2,
              r@3, 
              (r@4)^(r@8),
              (r@5)^(r@8),
              (r@6)^(r@8),
              r@7,
              r@8])
                where r = [False] # state.R)
            else (rList @i).R}
          | i <- [0..254]]
\end{verbatim}

Not only is this computationally slower since it runs 255 loops every time rc is called,
it differs vastly from the NIST specificiation. Using a new ‘while’ utility function, we 
can declare an initial state, condition function, and body function that will recursively 
be called until the condition is met. Here is the rc function with the use of this utility 
function:

\begin{verbatim}
  rc : [64] -> Bit
  rc t = if (t % 255) == 0 
    then True 
    else (while {i = 1, R = 128:[8]} //STATE
      (\state -> state.i <= t % 255) //COND
      (\state -> {                   //LOOP
        i = state.i + 1, 
        R = (take`{8} ([
          (r@0)^(r@8),
          r@1,
          r@2,
          r@3,
          (r@4)^(r@8),
          (r@5)^(r@8),
          (r@6)^(r@8),
          r@7, 
          r@8]) 
            where r = [False] # state.R)})
      ).R @0
\end{verbatim}

@@

\begin{verbatim}
  while : {a} a -> (a -> Bit) -> (a -> a) -> a
  while state cond f = if (cond state)
    then (while (f state) cond f)	
    else state
\end{verbatim}

While the condition is not met by the current state, continue to iterate. Else, return the 
current state.

>>>>

\section{Amended Cryptol}\label{sec:amended}
Initially it appeared that visual inspection with Cryptol’s standard syntax was subpar.
Cryptol’s functional folds and list comprehensions do not align easily with the imperative 
and array-based NIST standard. By leveraging Cryptol’s versatility, however, we produced a couple 
utility functions to visually mimic the sequential nature of the NIST specification.

The phrase “for all” indicates a looping structure and is quite common in the NIST standard. 
It is used in the definition of the pi function.

\begin{figure}[h]
  \centering
  \begin{verbatim}
1. For all triples (x,y,z) such that 
  0 <= x < 5, 0 <= y < 5, and 0 <= z < w,
  let A'[x, y, z] = A[(x + 3y) mod 5, x, z].
2. Return A'.
  \end{verbatim}
  \caption{NIST Pi definition}
  \label{fig:nistPi}
\end{figure}

Using Cryptol's list comprehensions, we find the following syntax for the pi function:

\begin{figure}[h]
  \centering
\begin{verbatim}
type COL = 5 
type W = 64
type STATE_ARR = [COL][COL][W]
LIST4 = [0..COL-1]
LIST63 = [0..W-1]
pi : STATE_ARR -> STATE_ARR
pi a = [[[a @x @((x + 3y) % COL) @z 
            | z <- LIST63] 
          | y <- LIST4] 
        | x <- LIST4]
\end{verbatim}
\caption{Cryptol Pi definition}
\label{fig:cryptolPi}
\end{figure}

@@

\begin{figure}[h]
  \centering
\begin{verbatim}
type COL = 5 
type W = 64
type STATE_ARR = [COL][COL][W]
LIST4 = [0..COL-1]
LIST63 = [0..W-1]
pi : STATE_ARR -> STATE_ARR
pi a = for LIST4 (\y ->
          for LIST4 (\x -> 
            for LIST4 (\z -> 
              a @x @((x + 3*y) % `COL) @z)))
\end{verbatim}
\caption{Ammended Cryptol Pi definition}
\label{fig:cryptolAmmendedPi}
\end{figure}

@@

\begin{figure}[h]
  \centering
\begin{verbatim}
for : {n, a, b} [n]a -> (a -> b) -> [n]b
for vals loop = [loop index | index <- vals]
\end{verbatim}
\caption{Cryptol For utility function}
\label{fig:cryptolFor}
\end{figure}

@@

\begin{figure}[h]
  \centering
\begin{verbatim}
1. If t mod 255 = 0, return 1.
2. Let R = 10000000.
3. For i from 1 to t mod 255, let:
  a. R = 0 || R;
  b. R[0] = R[0] ^ R[8];
  c. R[4] = R[4] ^ R[8];
  d. R[5] = R[5] ^ R[8];
  e. R[6] = R[6] ^ R[8];
  f. R =Trunc8[R].
4. Return R[0].
\end{verbatim}
\caption{NIST RC definition}
\label{fig:nistRC}
\end{figure}

@@

\begin{figure}[h]
  \centering
\begin{verbatim}
rc : [64] -> Bit
rc t = if (t % 255) == 0 
  then True 
  else rs !0 @0 where
    rs = [128:[8]] # [
      if i <= t % 255
        then (take`{8} ([
          (r@0)^(r@8),
          r@1,
          r@2,
          r@3, 
          (r@4)^(r@8),
          (r@5)^(r@8),
          (r@6)^(r@8),
          r@7,
          r@8])
            where r = [False] # (rs @(i - 1)))
        else rs @(i - 1)
      | i <- [1..254]]
\end{verbatim}
\caption{Cryptol RC definition}
\label{fig:cryptolRC}
\end{figure}

Not only is this computationally slower since it runs 255 loops every time rc is called, 
is stores the previous states in a list and indexes into it. This differs vastly from the 
NIST specificiation both visually and functionally. Using a new ‘while’ utility function, we 
can declare an initial state, condition function, and body function that will recursively 
be called until the condition is met. It consolodates the information of the previous state 
until the condition is sastified. Here is the rc function with the use of this utility 
function:

\begin{figure}[h]
  \centering
\begin{verbatim}
rc : [64] -> Bit
rc t = if (t % 255) == 0 
  then True 
  else (while {i = 1, R = 128:[8]} //STATE
    (\state -> state.i <= t % 255) //COND
    (\state -> {                   //LOOP
      i = state.i + 1, 
      R = (take`{8} ([
        (r@0)^(r@8),
        r@1,
        r@2,
        r@3,
        (r@4)^(r@8),
        (r@5)^(r@8),
        (r@6)^(r@8),
        r@7, 
        r@8]) 
          where r = [False] # state.R)})
    ).R @0
\end{verbatim}
\caption{Ammended Cryptol RC definition}
\label{fig:cryptolAmmendedRC}
\end{figure}

@@

\begin{figure}[h]
  \centering
\begin{verbatim}
while : {a} a -> (a -> Bit) -> (a -> a) -> a
while state cond f = if (cond state)
  then (while (f state) cond f)	
  else state
\end{verbatim}
\caption{Cryptol While utility function}
\label{fig:cryptolWhile}
\end{figure}
